---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# zoneR

<!-- badges: start -->
<!-- badges: end -->

Analyze the housing capacity of cities using Open Zoning Feed
Specification (OZFS) data.

The main function of zoneR is `zr_run_zoning_checks()` which reads in an
OZFS `.bldg`, `.parcel`, and `.zoning` file. It calculates the zoning
requirements for each parcel, checks them against the building
characteristics, and returns a data frame with a row for each parcel and
an "allowed" column containing one of three possible values.

- `TRUE`: The building is allowed on the parcel

- `FALSE`: The building is not allowed on the parcel

- `MAYBE`: The building might be allowed on the parcel depending on 
conditions that are not yet possible to express in OZFS. 

## Installation

You can install the development version of zoneR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("vibe-lab-gsd/zoneR")
```

## Example

This is a basic example of finding which parcels in Paradise, TX allow a duplex 
with a gross floor area of 3200 square feet and a height of 45 feet. 
Small OZFS files accessible through the package will be used in the example. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
```

We get the file paths for the example OZFS files using `zr_example_files()`.
```{r}
library(zoneR)

bldg_path <- zr_example_files("2_fam.bldg")
parcel_path <- zr_example_files("Paradise.parcel")
zoning_path <- zr_example_files("Paradise.zoning")
```

We can then put each of the files into the `zr_run_zoning_checks()` function and 
assign the result to a variable called zoning_checks. 
```{r}
zoning_checks <- zr_run_zoning_checks(bldg_file = bldg_path,
                                      parcel_files = parcel_path,
                                      zoning_files = zoning_path)
```

The result of the function is a simple features data frame with the following columns: "parcel_id", "allowed", "reason", "geometry." Note that the geometry is the centroid of the parcel. 
```{r}
head(zoning_checks)
```

The graph below is a visualization of the results. 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
parcels <- sf::st_read(parcel_path, quiet = T)

parcels <- parcels |>
  dplyr::filter(side != "centroid")  |>
  dplyr::group_by(parcel_id) |>
  dplyr::summarize(geometry = sf::st_polygonize(sf::st_union(geometry)))

parcel_polygons <- dplyr::left_join(parcels, sf::st_drop_geometry(zoning_checks), by = "parcel_id")

parcel_polygons |>
  ggplot() +
  geom_sf(aes(fill = allowed)) +
  scale_fill_manual(values = c("FALSE" = "#f4a582",
                               "MAYBE" = "#f7f7f7",
                               "TRUE" = "#92c5de"),
                    na.value = "grey70") +
  theme_void()
```

It appears this duplex would not be allowed in most the Paradise parcels.
To look further into the reasons, we can specify the checks we want to be run
and look at those results. Since land use regulations are often a distinguishing
factor in allowing a building on a parcel, only the res_type check will be run
in this next example. We will also put `print_checkpoints = FALSE` because we
don't really need it to tell us when it is done with the different parts of the
function. 

```{r}
zoning_checks <- zr_run_zoning_checks(bldg_file = bldg_path,
                                      parcel_files = parcel_path,
                                      zoning_files = zoning_path, 
                                      checks = "res_type",
                                      print_checkpoints = FALSE)
```

Now we can see that only a few of the parcels are in a district that allows
buildings with two units. It makes sense why there were only a few parcels 
in the previous check that might have allowed the building. 
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
parcels <- sf::st_read(parcel_path, quiet = T)

parcels <- parcels |>
  dplyr::filter(side != "centroid")  |>
  dplyr::group_by(parcel_id) |>
  dplyr::summarize(geometry = sf::st_polygonize(sf::st_union(geometry)))

parcel_polygons <- dplyr::left_join(parcels, sf::st_drop_geometry(zoning_checks), by = "parcel_id")

parcel_polygons |>
  ggplot() +
  geom_sf(aes(fill = allowed)) +
  scale_fill_manual(values = c("FALSE" = "#f4a582",
                               "MAYBE" = "#f7f7f7",
                               "TRUE" = "#92c5de"),
                    na.value = "grey70") +
  theme_void()
```

Here are a few other constraint checks to visualize. 

Height Check
```{r}
zoning_checks <- zr_run_zoning_checks(bldg_file = bldg_path,
                                      parcel_files = parcel_path,
                                      zoning_files = zoning_path, 
                                      checks = "height",
                                      print_checkpoints = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
parcels <- sf::st_read(parcel_path, quiet = T)

parcels <- parcels |>
  dplyr::filter(side != "centroid")  |>
  dplyr::group_by(parcel_id) |>
  dplyr::summarize(geometry = sf::st_polygonize(sf::st_union(geometry)))

parcel_polygons <- dplyr::left_join(parcels, sf::st_drop_geometry(zoning_checks), by = "parcel_id")

parcel_polygons |>
  ggplot() +
  geom_sf(aes(fill = allowed)) +
  scale_fill_manual(values = c("FALSE" = "#f4a582",
                               "MAYBE" = "#f7f7f7",
                               "TRUE" = "#92c5de"),
                    na.value = "grey70") +
  theme_void()
```

Building Fit Check
```{r}
zoning_checks <- zr_run_zoning_checks(bldg_file = bldg_path,
                                      parcel_files = parcel_path,
                                      zoning_files = zoning_path, 
                                      checks = "bldg_fit",
                                      print_checkpoints = FALSE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
parcels <- sf::st_read(parcel_path, quiet = T)

parcels <- parcels |>
  dplyr::filter(side != "centroid")  |>
  dplyr::group_by(parcel_id) |>
  dplyr::summarize(geometry = sf::st_polygonize(sf::st_union(geometry)))

parcel_polygons <- dplyr::left_join(parcels, sf::st_drop_geometry(zoning_checks), by = "parcel_id")

parcel_polygons |>
  ggplot() +
  geom_sf(aes(fill = allowed)) +
  scale_fill_manual(values = c("FALSE" = "#f4a582",
                               "MAYBE" = "#f7f7f7",
                               "TRUE" = "#92c5de"),
                    na.value = "grey70") +
  theme_void()
```

Many of the "MAYBE" values attached to the bldg_fit constraint represent parcels that didn't 
have labeled sides. Without labeled sides, a proper buildable area could not be calculated,
and the building fit check was skipped. 

#### Detailed Check

In addition to the usual data frame that is returned, setting `detailed_check = TRUE` 
returns the data frame with additional columns for every check that took place 
(Note this requires more processing time as the code does not skip 
consecutive checks if a previous check yields FALSE).


```{r}
zoning_checks <- zr_run_zoning_checks(bldg_file = bldg_path,
                                      parcel_files = parcel_path,
                                      zoning_files = zoning_path,
                                      detailed_check = TRUE)
```

With these added columns, we can look more closely at why the building does not fit in
most of the parcels.

```{r, results='hide'}
column_true_sums <- colSums(sf::st_drop_geometry(zoning_checks) == "TRUE", na.rm = TRUE)

ordered_sums <- column_true_sums[order(column_true_sums)]
```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
sums_df <- data.frame(constraint_check = names(ordered_sums), parcels_true = ordered_sums) |>
  dplyr::filter(!constraint_check %in% c("parcel_id","allowed","reason"))

row.names(sums_df) <- seq(nrow(sums_df))

knitr::kable(sums_df)
```


It appears the res_type constraint is the most restrictive and the unit_size 
constraint is the least restrictive for the given building. 

## Inside the zr_run_zoning_checks()
Most of the functions in zoneR were created for `zr_run_zoning_checks()`, but
they can be use by themselves if desired. Below are some of the main variables that 
the different zoneR functions take in.

-   **bldg_data:** The `.bldg` file read in as a list using
    `rjson::fromJSON()`

-   **parcel_dims:** Created with the `zr_get_parcel_dims()` function. A
    simple features data frame with all the centroid and dimensional
    data from the `.parcel` file. It contains one row per parcel.

-   **parcel_data:** One row of the parcel_dims data frame representing
    a unique parcel

-   **parcel_geo:** Created with the `zr_get_parcel_geo()` function. A
    simple features data frame containing the geometry of each parcel
    side without the centroid or dimensional data.

-   **zoning_data:** The data in the `.zoning` file read in as a simple
    features data frame using `sf::st_read()`.

-   **district_data:** One row of the zoning_data data frame
    representing a unique district.
